Machine Learning guided workflow for Ribosome Binding Site engineering

### 2021-05-24 paper/main.tex
Updates from Overleaf

One of the main tenets of synthetic biology is design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory. 
Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used \cite{Yeoh2019,Nielsen2016}.
Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements. 
This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.

The biggest limitation for the DBTL approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
Reported predictions from all of these models are relatively good ($R^2 >0.8$), 
% \mengyan{Maciej: reviewer might wonder why they got high R2 but we don't, e.g. is that because the data? So I recommend either remove this or explain it somewhere.} \maciej{I'll explain it in the discussion}
but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}. 
There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980} . This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016} .

Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
These new datasets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020}. 
In the past few years there was a significant uptake of machine learning based approaches in synthetic biology \cite{LAWSON2021}.
Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}.
There were also successful attempts at using deep learning techniques for analysis of big datasets \cite{Alipanahi2015,Angermueller2016}.

% Machine learning has been also used for prediction in proteins \cite{Yang2018}.
However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. 
The challenges of applying machine learning techniques on synthetic biology have two folds.
On the one hand, there is no large-scale and high-quality labelled data. 
The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
And it is time-consuming and expensive to measure the TIR for RBS sequences. \mengyan{Maciej: can you add more details or reference here}.
One the other hand,  while the goal is to find highest TIR among the design space, one also need to explore as much as possible to learn the whole design space to avoid getting stuck in the local maximum.  
We addressed the challenges by applying DBTL workflow. 
Instead of querying labels for all RBS sequences in the design space (4,138 in total), we recommend RBS sequences to query in batches and learn the prediction model in an online form.
Particularly, we use bandits algorithm for recommending RBS sequences to address the exploitation-exploration balance. 
% For the machine learning to work effectively the analysed data set needs to show two qualities: high relative volume and high quality of data.
% These two elements don't have specific definitions, but in general the data set has to be going into at least hundreds data points and have to cover 5-10\% of that space.
% \mengyan{Maciej: I think it varies from case to case. We need to either change it or add reference.}
% And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\

We present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with the goal of optimising associated protein expression level. 
RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
In this work we have used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS .
Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by querying the batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.
We did this by designing a sequential experimental workflow shown in Figure \ref{fig: Flowchart}.
In the zeroth round, randomised RBS sequences and preliminary machine learning recommended designs based on literature were designed to explore the experimental space. 
In the subsequent rounds, designs were recommended by the algorithm based on the data obtained in the previous rounds. 
The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
After constructing, the plasmids harbouring the new genetic devices were tested in microplate reader.
The results were then fed back to the algorithm for it to recommend the next round of designs.
This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.
At the same time we were able to discover new RBS sequences with very high TIRs from between 95 to 135\% of TIR of our chosen very strong benchmark RBS.

@@

We show our DBTL workflow in Figure \ref{fig: Flowchart}.
BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
Machine learning algorithms are applied in LEARN and DESIGN.
In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
The goal is to provide predictions about TIR labels and uncertainty to assist the recommendation process.
In the DESIGN phase, we have used a multi-armed Bandit recommendation algorithm that was selecting RBS designs to be tested in subsequent experimental batches.
\\

@@

In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. 
By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring plasmid with the device over time.
Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.

In \emph{E. coli}, the RBS is usually located in the 20 bases upstream of the start codon. 
Additionally, there is a consensus RBS core sequence called the \textit{Shine-Dalgarno sequence}, which in \emph{E. coli} is \textbf{AGGAGG}. 
Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region
(see detail in Figure \ref{fig: Flowchart}).

Our template RBS sequence, is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA.
This sequence is a known to have high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
% Due to poor testing methods that were used for TIR previously we can't claim "highest" cause we don't know how do we compare exactly
Since this is the sequence against which new RBS sequenced will be benchmarked,
we will refer to this sequence as the \textit{benchmark sequence} from now on.
In our design we focus on randomising of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the benchmark sequence, i.e. TTTAAGA + NNNNNN + TATACAT, where N can be any choices of A, C, G, T. 
The total experimental (variant) space is then $4^6$ = 4096.
We have experimentally confirmed that changing the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).

@@

Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
% Different randomly generated groups were run together with Bandit 0 (based on literature data) group in Round 0.
% Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
In each round, we measure the TIR of benchmark RBS as the internal standard. 
We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).
All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
The Bandit-0 group performed poorly, despite being machine learning driven, due to being trained on literature data.
However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.
Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms' better understanding of the experimental space with more data.
The decreased result in the 0.9th qunatile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
We see that effect in Figure \ref{fig: Swarmplot and Quantplot} c), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
We can see the RBSs with high TIRs (blue, with high predicted mean) tend to come from exploitation the design space whereas the explorative points (red, with high predicted standard deviation) give relatively low TIR but expand our knowledge about the unknown part of the design space. 
Figure \ref{fig: Swarmplot and Quantplot}D shows the TIRs of RBSs tested in the Bandit groups divided into bins of width of 0.1 TIR ratio.
KDE plots have been overlaid to depict the calculated density for each group.
The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the $>0.8$ TIR ratio bins.
\mengyan{do we want to address two-mode pattern (due to the exploration and exploitation.)}
\maciej{I did that in a later section.}
\mengyan{I recommend to mention the observation and point to where you discuss it.}
In \ref{fig: Swarmplot and Quantplot}E we show a tSNE plot depicting the experimental space.
Each RBS is located on the plot according to its distance to other RBSs as calculated by our kernel function.
We can see the RBSs recommended by Bandit groups have covered majority of the design space. 
A number of clusters were targeted by our recommendation algorithm.
For example, the circled clusters labelled as "G-Rich Clusters" have been actively recommended by the algorithm.
More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.

@@

Our recommendation algorithm selects new designs based on means and uncertainties calculated by our prediction algorithm.
This algorithm creates a model which takes the RBS sequences as input and predicts the TIR values with uncertainty level about the prediction, based on the experimental data.
For this study, we have used the Gaussian Process regression (GPR) as our prediction method.
GPR is a Bayesian approach and has been widely used for experimental design \cite{srinivas2012information, romero_navigating_2013}.
% Such a stochastic prediction method fits biological processes naturally, since they are highly stochastic as well. 
The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.\\
A crucial ingredient in a Gaussian Process predictor \cite{Rasmussen2004} is the kernel (covariance) function, which captures the similarity between data points, in our case RBS sequences.
Specifically, kernel function implicitly embeds RBS sequences into high-dimensional feature space which makes the regression process easier.
For Bandit designs in Round 0, since we only had access to limited number of data points from literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
For subsequent rounds, we used 
% a more powerful \maciej{more powerful in what sense?} kernel function, 
the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds}, which has been shown to have good performance in various prediction tasks \cite{Ben-Hur2008}.
The WDS not only counts the matches of substrings of a certain length (i.e. kmers), but also takes into account the positional information and shifting of substrings.
For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift.
The relative distances between sequences calculated by our kernel function are shown in figure \ref{fig: Swarmplot and Quantplot}E.\\

Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
As expected, the predictions in Round 0 were poor due to use of approximated data. 
The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.065 for round 0 to R\textsuperscript{2} of 0.27 for round 3.
Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.
% Interestingly there wasn't much difference in fitness parameters between Round 1-3.
% One interesting point is that despite relatively low correlations between predicted and true values our recommendation algorithm performed very well.
One important point to note is that the predictions are also influenced by our recommendation choices. 
In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having real mean different to what was predicted.
However, this is still very useful information for future predictions as it allows us to understand the underlying space better.

@@

Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 (in terms of TIR ration) of our sequences.
It is generally understood that guanine rich sequences are promoting strong transcription.
This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
This result combined with the Bandits' algorithm bias towards the G rich cluster shown in figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G rich sequences as ones with high TIR probability.

@@

\begin{table}[!h]
\centering
\begin{minipage}[c]{0.6\textwidth}
\centering
\begin{tabular}{|c|c|}
\hline
\textbf{Characteristics of the library}                                                       & \textbf{Statistics} \\ \hline
Total experimental space                                                                      & 4096                \\ \hline
Planned constructs                                                                            & 450                 \\ \hline
Successfully constructed                                                                      & 445                 \\ \hline
Sequences with CV\textless{}40\%                                                              & 79\%                \\ \hline
Sequences with CV\textless{}20\%                                                              & 27\%                \\ \hline
\begin{tabular}[c]{@{}c@{}}Efficiency of bandit design \\ (compared with random)\end{tabular} & 2                   \\ \hline
Raw TIR range                                                                                     &      [4.9258, 105.377]               \\ \hline
TIR ratio range                                                                                     &      [0.0626, 1.3393]               \\ \hline
\end{tabular}
\end{minipage}
\begin{minipage}[c]{0.38\textwidth}
\centering
% Please add the following required packages to your document preamble:
% \usepackage[table,xcdraw]{xcolor}
% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
\begin{tabular}{|c|c|}
\hline
\textbf{Top RBS Core} & \textbf{TIR Ratio} \\ \hline
GGGGGC                & 1.3393             \\ \hline
GGGGGT                & 1.1544             \\ \hline
GGCTAT                & 1.084              \\ \hline
\textbf{AGGAGA}                & 1                  \\ \hline
GGCGTT                & 0.9805             \\ \hline
GGGGGG                & 0.9788             \\ \hline
GGCGAC                & 0.9758             \\ \hline
CAGGAG                & 0.9629             \\ \hline
GGCGAG                & 0.9519             \\ \hline
\textbf{AGGAGG}                & 0.3941             \\ \hline
\end{tabular}
\end{minipage}
\caption{Characteristics of the library. \mengyan{write caption here}
For Efficiency of bandit design:
random = {set of TIR that are in UNI and PPM};
bandit = {set of TIR that are in bandit 0,1,2,3}.
The ratio of the max bandit to the max random: $1.34/0.67$ = 2;
The ratio of the 0.9 quantile bandit to the 0.9 quantile random: $0.82/0.49$ = 1.67.}
\end{table}

@@

\mengyan{We should try to alter the discussion in terms of our three-sentence story, along with open questions (in a more positive way) and future work.}

>>>>

One of the main tenets of synthetic biology is design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory. 
Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used \cite{Yeoh2019,Nielsen2016}.
Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements. 
This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.\\
The biggest limitation for the DBTL approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
Reported predictions from all of these models are relatively good ($R^2 >0.8$), 
but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}. 
There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}. This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.\\
Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
These new datasets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020}. 
In the past few years there was a significant uptake of machine learning based approaches in synthetic biology \cite{LAWSON2021}.
Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}.
There were also successful attempts at using deep learning techniques for analysis of big datasets \cite{Alipanahi2015,Angermueller2016}.\\
However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. 
The challenges of applying machine learning techniques on synthetic biology have two folds.
On the one hand, there is no large-scale and high-quality labelled data. 
The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
And it is time-consuming and expensive to measure the TIR for RBS sequences. 
On the other hand,  while the goal is to find highest TIR among the design space, one also need to explore as much as possible to learn the whole design space to avoid getting stuck in the local maximum.  
We addressed the challenges by applying DBTL workflow. 
Instead of querying labels for all RBS sequences in the design space (4,138 in total), we recommend RBS sequences to query in batches and learn the prediction model in an online form.
Particularly, we use bandits algorithm for recommending RBS sequences to address the exploitation-exploration balance. \\
We present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with the goal of optimising associated protein expression level. 
RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
In this work we have used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS .
Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by querying the batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.
We did this by designing a sequential experimental workflow shown in Figure \ref{fig: Flowchart}.
In the zeroth round, randomised RBS sequences and preliminary machine learning recommended designs based on literature were designed to explore the experimental space. 
In the subsequent rounds, designs were recommended by the algorithm based on the data obtained in the previous rounds. 
The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
After constructing, the plasmids harbouring the new genetic devices were tested in microplate reader.
The results were then fed back to the algorithm for it to recommend the next round of designs.
This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.
At the same time we were able to discover new RBS sequences with very high TIRs from between 95 to 135\% of TIR of our chosen very strong benchmark RBS.

@@

We show our DBTL workflow in Figure \ref{fig: Flowchart}.
BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
Machine learning algorithms are applied in LEARN and DESIGN.
In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
The goal is to provide predictions about TIR labels and uncertainty to assist the recommendation process.
In the DESIGN phase, we have used a multi-armed Bandit recommendation algorithm that was selecting RBS designs to be tested in subsequent experimental batches.\\

@@

In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. 
By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring plasmid with the device over time.
Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.\\
In \emph{E. coli}, the RBS is usually located in the 20 bases upstream of the start codon. 
Additionally, there is a consensus RBS core sequence called the \textit{Shine-Dalgarno sequence}, which in \emph{E. coli} is \textbf{AGGAGG}. 
Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region
(see detail in Figure \ref{fig: Flowchart}).\\
Our template RBS sequence, is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA.
This sequence is a known to have high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
Since this is the sequence against which new RBS sequenced will be benchmarked,
we will refer to this sequence as the \textit{benchmark sequence} from now on.
In our design we focus on randomising of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the benchmark sequence, i.e. TTTAAGA + NNNNNN + TATACAT, where N can be any choices of A, C, G, T. 
The total experimental (variant) space is then $4^6$ = 4096.
We have experimentally confirmed that changing the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).

@@

Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
In each round, we measure the TIR of benchmark RBS as the internal standard. 
We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).
All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
The Bandit-0 group performed poorly, despite being machine learning driven, due to being trained on literature data.
However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.
Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms' better understanding of the experimental space with more data.
The decreased result in the 0.9th qunatile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
We see that effect in Figure \ref{fig: Swarmplot and Quantplot} c), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
We can see the RBSs with high TIRs (blue, with high predicted mean) tend to come from exploitation the design space whereas the explorative points (red, with high predicted standard deviation) give relatively low TIR but expand our knowledge about the unknown part of the design space. 
Figure \ref{fig: Swarmplot and Quantplot}D shows the TIRs of RBSs tested in the Bandit groups divided into bins of width of 0.1 TIR ratio.
KDE plots have been overlaid to depict the calculated density for each group.
The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the $>0.8$ TIR ratio bins.
Notably, the distributions calculated for all the groups show bivariate distribution, we discuss the possible reasons for that further in the text.\\
In \ref{fig: Swarmplot and Quantplot}E we show a tSNE plot depicting the experimental space.
Each RBS is located on the plot according to its distance to other RBSs as calculated by our kernel function.
We can see the RBSs recommended by Bandit groups have covered majority of the design space. 
A number of clusters were targeted by our recommendation algorithm.
For example, the circled clusters labelled as "G-Rich Clusters" have been actively recommended by the algorithm.
More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.

@@

Our recommendation algorithm selects new designs based on means and uncertainties calculated by our prediction algorithm.
This algorithm creates a model which takes the RBS sequences as input and predicts the TIR values with uncertainty level about the prediction, based on the experimental data.
For this study, we have used the Gaussian Process regression (GPR) as our prediction method.
GPR is a Bayesian approach and has been widely used for experimental design \cite{srinivas2012information, romero_navigating_2013}.
The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.\\
A crucial ingredient in a Gaussian Process predictor \cite{Rasmussen2004} is the kernel (covariance) function, which captures the similarity between data points, in our case RBS sequences.
Specifically, kernel function implicitly embeds RBS sequences into high-dimensional feature space which makes the regression process easier.
For Bandit designs in Round 0, since we only had access to limited number of data points from literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
For subsequent rounds, we used 
% a more powerful \maciej{more powerful in what sense?} kernel function, 
the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds}, which has been shown to have good performance in various prediction tasks \cite{Ben-Hur2008}.
The WDS not only counts the matches of substrings of a certain length (i.e. kmers), but also takes into account the positional information and shifting of substrings.
For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift.
The relative distances between sequences calculated by our kernel function are shown in figure \ref{fig: Swarmplot and Quantplot}E.\\

Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
As expected, the predictions in Round 0 were poor due to use of approximated data. 
The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.065 for round 0 to R\textsuperscript{2} of 0.27 for round 3.
Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.
One important point to note is that the predictions are also influenced by our recommendation choices. 
In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having real mean different to what was predicted.
However, this is still very useful information for future predictions as it allows us to understand the underlying space better.

@@

Our data set taken together can be viewed as a reliable library of RBS sequences for \emph{E. coli}, some characteristics of which are shown in table 1.
Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 (in terms of TIR ration) of our sequences.
It is generally understood that guanine rich sequences are promoting strong transcription.
This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
This result combined with the Bandits' algorithm bias towards the G rich cluster shown in figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G rich sequences as ones with high TIR probability.

@@

\begin{table}[!h]
\centering
\begin{minipage}[c]{0.6\textwidth}
\centering
\begin{tabular}{|c|c|}
\hline
\textbf{Characteristics of the library}                                                       & \textbf{Statistics} \\ \hline
Total experimental space                                                                      & 4096                \\ \hline
Planned constructs                                                                            & 450                 \\ \hline
Successfully constructed                                                                      & 445                 \\ \hline
Sequences with CV\textless{}40\%                                                              & 79\%                \\ \hline
Sequences with CV\textless{}20\%                                                              & 27\%                \\ \hline
\begin{tabular}[c]{@{}c@{}}Efficiency of bandit design \\ (compared with random)\end{tabular} & 2                   \\ \hline
Raw TIR range                                                                                     &      [4.9258, 105.377]               \\ \hline
TIR ratio range                                                                                     &      [0.0626, 1.3393]               \\ \hline
\end{tabular}
\end{minipage}
\begin{minipage}[c]{0.38\textwidth}
\centering
% Please add the following required packages to your document preamble:
% \usepackage[table,xcdraw]{xcolor}
% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
\begin{tabular}{|c|c|}
\hline
\textbf{Top RBS Core} & \textbf{TIR Ratio} \\ \hline
GGGGGC                & 1.3393             \\ \hline
GGGGGT                & 1.1544             \\ \hline
GGCTAT                & 1.084              \\ \hline
\textbf{AGGAGA}                & 1                  \\ \hline
GGCGTT                & 0.9805             \\ \hline
GGGGGG                & 0.9788             \\ \hline
GGCGAC                & 0.9758             \\ \hline
CAGGAG                & 0.9629             \\ \hline
GGCGAG                & 0.9519             \\ \hline
\textbf{AGGAGG}                & 0.3941             \\ \hline
\end{tabular}
\end{minipage}
\caption{\textbf{Characteristics of the library.}
Left table presents some of the characteristics of our library.
Right table presents 10 RBS sequences with their corresponding TIR ratios, the first 9 being the strongest sequences including the benchmark sequence (AGGAGA) and the last being the Shine-Dalgarno sequence (AGGAGG).
CV is coefficient of variation (STD of a sample divided by its mean).
Efficiency of the bandit design is calculated by dividing the highest TIR found using machine learning by highest TIR found using random sequence generation. }
\end{table}

@@
